SIR-RL: Reinforcement Learning for Optimized Policy Control during Epidemiological Outbreaks in Emerging Market and Developing Economies

### 2024-03-06 paper_draft/main.tex
recompiled paper
paper edits: fig captions, sirv to sir with v

\title{SIRV-RL: Reinforcement Learning for Optimized Policy Control during Epidemiological Outbreaks in Emerging Market and Developing Economies}
\author[1,*]{Maeghal Jain}
\author[1]{Ziya Uddin}

@@

In this paper we optimize the government policies regarding stringency as it controls both the spread of the disease and the economy. To model the epidemiological data~\cite{WorldometerCorona} we use the simple SIR model without vital dynamics~\cite{Hethcote1989, Hethcote2008, ALLEN2017128} as it is assumed that the timescale is small enough that it can be neglected~\cite{Cooper2020}. By lesioning the model, as opposed to proposing a new mathematical model with more specialized compartments to more accurately represent the actual environment~\cite{Bjrnstad2020, Mwalili2020}, we effectively model the disease progression. Our model (SIRV with lockdown and time-varying vaccination rate) builds on the foundational SIR model, by accounting for the recovery reached through vaccination~\cite{Marinov2022,MaurciodeCarvalho2023,Thater,Turkyilmazoglu2022,YALADANDA2022101052} and the effects of lockdown~\cite{Hale2021,SIRLockdown,NBERw26981,NBERw26867}. Although the traditional SIR model is a valuable tool for understanding the spread of infectious diseases, it assumes that parameters like the transmission rate ($\beta$) and recovery rate ($\gamma$) are constant over time, which may not always be the case. In this paper, we propose a more sophisticated approach by introducing a time-dependent SIR model~\cite{TimeDependentSIR}, enabling us to account for the changing dynamics of the pandemic due to factors such as lockdowns and vaccination rates. This proposed model effectively addresses the real-world conditions acts as a solution that is both effective and extendable. However, the study has limitations; First, the deterministic SIR model (predecessor to our proposed model) fails to account for chance in disease spread and lacks confidence intervals on results and while stochastic models incorporate chance, they are typically more challenging to analyze than their deterministic counterparts~\cite{Hethcote2008}. Second, the underreporting of cases during the period selected by our study. Lastly, the reinforcement learning agent should be resistant to how the vaccination rate changes -- or to when it is first introduced because then it scopes the environment for succumbing to wishful thinking which can be  potentially dangerous.

@@

And, $\beta$ the effective transmission rate is defined as:
\begin{equation}
  \beta = p c
  \label{eq:beta_effective_transmission_rate}
\end{equation}
During an epidemic, the fundamental drivers of an epidemic growth is the rate of infection $\beta$ i.e. the average number of infections per infected case and the infectious period $1/\gamma$ i.e. the average period for which the infected case is infected for. Epidemics can only happen if the case is infectious enough for long enough and this defined by $R_0 = \beta / \gamma$. Here, $R_0$ is The average number of secondary infections caused by each infected case, in an otherwise fully susceptible population.

@@

To estimate the parameters $\beta$ and $\gamma$ for India, based on the data from May, 2020, to October, 2022, we simply define two cost functions \cref{eq:cost_SIR,eq:cost_I} to calibrate the model with the use of huber loss~\cite{huberloss}. We generally calibrate the model using \cref{eq:cost_SIR}. However, sometimes we must strike a balance between modeling all three population groups (susceptible, infected, and recovered) and focusing just on the infected group, which is most important because it drives the spread of the disease, therefore, in that case we consider the loss obtained from \cref{eq:cost_SIR,eq:cost_I} (see \cref{optimizing_window_length}, where a hyperparameter is selected keeping both these losses in mind).
\begin{equation}
  L_{\delta}(y, f(x)) = 
  \begin{cases}
    \frac{1}{2}{(y - f(x))^2} & \text{for } |y - f(x)| \leq \delta, \\
    \delta \cdot (|y - f(x)| - \frac{1}{2}\delta) & \text{otherwise.} \\
  \end{cases}
  \label{eq:huberloss}
\end{equation}
In the above equation, $y$ is the actual data and $f(x)$ is the prediction.
\begin{equation}
  \begin{split}
  \textrm{loss\_SIR} = \textrm{cost\_function\_SIR}(S, \hat{S}, I, \hat{I}, R, \hat{R}) \\
  = L_{\delta = 1}(S, \hat{S}) + L_{\delta = 1}(I, \hat{I}) + L_{\delta = 1}(R, \hat{R})
  \end{split}
  \label{eq:cost_SIR}
\end{equation}
\begin{equation}
  \textrm{loss\_I} = \textrm{cost\_function\_I}(I, \hat{I}) = L_{\delta = 1}(I, \hat{I})
  \label{eq:cost_I}
\end{equation}
Where, $S$ is the number of susceptible people and $\hat{S}$ is the predicted number of susceptible people, similarly, $I$ for infected and $\hat{I}$ for the predicted number of infected people, and $R$ for recovered. Using the equations \Crefrange{eq:S_without_lockdown}{eq:R_without_lockdown}, we minimize the cost function \cref{eq:cost_SIR_without_lockdown} using the Nelder-Mead method\cite{Gao2010} to estimate the parameters like $\beta$, $\gamma$, and fit the model to actual data. The following parameters and loss is obtained:
\begin{equation}
  \beta_{optimal} = 0.042
  \label{eq:beta_optimal_without_lockdown}
\end{equation}
\begin{equation}
  \gamma_{optimal} = 0.024
  \label{eq:gamma_optimal_without_lockdown}
\end{equation}
\begin{equation}
  R_{0_{optimal}} = \frac{\beta_{optimal}}{\gamma_{optimal}} = 1.762
  \label{eq:r0_without_lockdown}
\end{equation}
\begin{equation}
  \textrm{loss\_SIR} = 85051490.533
  \label{eq:cost_SIR_without_lockdown}
\end{equation}
\begin{equation}
  \textrm{loss\_I} = 45187665.281
  \label{eq:cost_I_without_lockdown}
\end{equation}
See \cref{fig:SIR_model_IND} to see how the model compares with the actual data.

\subsection{SIR Model with Lockdown}
Now that, a simple SIR model has been established -- we need to model the effects of the stringency index (measure for the strictness of lockdown) on $\beta$ (the effective transmission rate). To do this, we say the flow of susceptibles not only depend on $\beta$ but also $s(t)$ the stringency index at time~\cite{NBERw26867,AlvarezLockdownSIR,SIRLockdown,SIRLockdown2,analytical_covid_lockdown_model}. The stringency index is a composite measure based on nine response indicators including school closures, workplace closures, and travel bans, rescaled to a value from 0 to 100 (100 = strictest)~\cite{owidcoronavirus}. This index simply records the strictness of government policies and does not measure or imply the appropriateness or effectiveness of a country's response i.e. a higher score does not necessarily mean that a country's response is "better" than others lower on the index. 
To define the new time-varying beta that is dependent on the current stringency index, the following equations have been formulated:
\begin{equation}
  \frac{dS}{dt} = -\beta (1 - s(t)/100) \frac{S I}{N}
  \label{eq:S_with_lockdown}
\end{equation}
\begin{equation}
  \frac{dI}{dt} = \beta (1 - s(t)/100) \frac{S I}{N} - \gamma I
  \label{eq:I_with_lockdown}
\end{equation}
\begin{equation}
  \frac{dR}{dt} = \gamma I
  \label{eq:R_with_lockdown}
\end{equation}
Where, $s(t)$ is the stringency index at time $t$ and is scaled down by a factor of 100 to normalize it and bring it in the range $s(t)/100 \in [0, 1]$. Multiplying the rate of flow from $S$ to $I$ compartment with $1 - s(t)/100$ allows us to account for the effect that stringency has on the disease progression. A higher stringency index can theoretically, stop the flow from the susceptible population to the infected population entirely. Optimizing these equations with \cref{eq:cost_SIR} using the Nelder-Mead method, we get the following parameters and loss:
\begin{equation}
  \beta_{optimal} = 0.401
  \label{eq:beta_optimal_with_lockdown}
\end{equation}
\begin{equation}
  \gamma_{optimal} = 0.090
  \label{eq:gamma_optimal_with_lockdown}
\end{equation}
\begin{equation}
  \begin{split}
    \overbar{R_0} = 1.693 \quad \text{(Mean)} \\
    \widetilde{R_0} = 1.624 \quad \text{(Median)} \\
    \text{Mode}(R_0) = 0.804 \quad \text{(Mode)} \\
    \sigma_{R_0} = 0.786 \quad \text{(Standard Deviation)} \\
    R_0 \in [0.16467, 3.0497] \quad \text{(Range)}
  \end{split}
  \label{eq:r0_with_lockdown}
\end{equation}
\begin{equation}
  \textrm{loss\_SIR} = 98438821.456
  \label{eq:cost_SIR_with_lockdown}
\end{equation}
\begin{equation}
  \textrm{loss\_I} = 11345389.686
  \label{eq:cost_I_with_lockdown}
\end{equation}
\subsection{SIRV Model with Lockdown}
Lastly, an additional flow from the susceptible to recovered population can be shown by adding a vaccination rate $\nu$ in the model. 
\begin{equation}
  \frac{dS}{dt} = -\beta (1 - s(t)/100) \frac{S I}{N} - \nu S
  \label{eq:S_with_lockdown_and_nu}
\end{equation}
\begin{equation}
  \frac{dI}{dt} = \beta (1 - s(t)/100) \frac{S I}{N} - \gamma I
  \label{eq:I_with_lockdown_and_nu}
\end{equation}
\begin{equation}
  \frac{dR}{dt} = \gamma I + \nu S
  \label{eq:R_with_lockdown_and_nu}
\end{equation}
Optimizing these equations with \cref{eq:cost_SIR} using the Nelder-Mead method:
\begin{equation}
  \beta_{optimal} = 0.409
  \label{eq:beta_optimal_with_lockdown_and_nu}
\end{equation}
\begin{equation}
  \gamma_{optimal} = 0.092
  \label{eq:gamma_optimal_with_lockdown_and_nu}
\end{equation}
\begin{equation}
  \nu_{optimal} = 2.904 \times 10^{-5}
  \label{eq:nu_optimal_with_lockdown_and_nu}
\end{equation}
\begin{equation}
  \begin{split}
    \overbar{R_0} = 1.691 \\
    \widetilde{R_0} = 1.623 \\
    \text{Mode}(R_0) = 0.803 \\
    \sigma_{R_0} = 0.785 \\
    R_0 \in [0.165, 3.047]
  \end{split}
  \label{eq:r0_with_lockdown_and_nu}
\end{equation}
\begin{equation}
  \textrm{loss\_SIR} = 94636860.384
  \label{eq:cost_SIR_with_lockdown_and_nu}
\end{equation}
\begin{equation}
  \textrm{loss\_I} = 10840360.995
  \label{eq:cost_I_with_lockdown_and_nu}
\end{equation}

\subsection{Optimizing Window Length for Time-varying Vaccination Rate}\label{optimizing_window_length}
However, as observed by the value of $\nu_{optimal}$ from \cref{eq:nu_optimal_with_lockdown_and_nu} which is almost negligible and the overestimation of infected individuals in \cref{fig:SIRV_model_with_lockdown_infections_IND} suggests that $\nu$ might be varying with time. This suggests to accurately estimate the infected population a time-varying vaccination rate should be used as the transition from susceptibility to direct recovery fluctuates with time~\cite{Liang2021,Marinov2022}. Therefore, using the $\beta_{optimal}$ and $\gamma_{optimal}$ from \cref{eq:beta_optimal_with_lockdown_and_nu} and \cref{eq:gamma_optimal_with_lockdown_and_nu}, we first find the optimal window length~\cite{Liao2020} i.e., a window length that results in the least loss from \cref{eq:cost_SIR,eq:cost_I}. Using different window lengths ($\textrm{window\_length} = 5, 10, 15 \dots 40, 45, 50$), we estimate value of $\nu$ given a sub-interval of $[{start}, {start} + {window\_length}]$ then for the next time-step $[{start} + {window\_length}, {start} + {window\_length} + {window\_length}]$ we again estimate $\nu$ -- so on and so forth. It is crucial to note that the variable $\nu$ is constrained to be a positive integer, reflecting the inherent one-way nature of vaccination: individuals can only receive vaccinations, not return them.
\begin{equation}
  \frac{dS}{dt} = -\beta_{optimal}  (1 - s(t)/100)  \frac{S I}{N} - \nu S
  \label{eq:S_with_lockdown_and_nu_calc_nu}
\end{equation}
\begin{equation}
  \frac{dI}{dt} = \beta_{optimal}  (1 - s(t)/100) \frac{S I}{N} - \gamma_{optimal} I
  \label{eq:I_with_lockdown_and_nu_calc_nu}
\end{equation}
\begin{equation}
  \frac{dR}{dt} = \gamma_{optimal} I + \nu S
  \label{eq:R_with_lockdown_and_nu_calc_nu}
\end{equation}
This model gives shows how $\nu$ varies with time. Since, the window length acts like a hyperparameter, we can try different window lengths in order to evaluate which would give us the best results. Therefore, the below graphs show us the loss using \cref{eq:cost_SIR,eq:cost_I}.

\begin{figure}[htbp!]
  \caption{Loss for Different Window Lengths}
  \begin{subfigure}[t]{0.48\textwidth}
    \centering
    \includegraphics[width=\linewidth]{images/window_length_loss_SIR_IND.pdf}
    \caption{Loss for Different Window Lengths for Susceptible, Infected and Recovered Population}
    \label{fig:window_length_loss_SIR_IND}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.48\textwidth}
    \centering
    \includegraphics[width=\linewidth]{images/window_length_loss_I_IND.pdf}
    \caption{Loss for Different Window Lengths for Infected Population}
    \label{fig:window_length_loss_I_IND}
  \end{subfigure}
\end{figure}

The following graph indicates that a window length of $10$ days yields the least overall loss for all three population groups. However, this window length results in a poor approximation for the infected group, which is crucial for accurately modeling the spread of the disease. Consequently, we have decided to use a window length of $15$ days, which provides a more accurate approximation for the infected population while still maintaining reasonable loss for the other groups.

\begin{figure}[htbp!]
  \centering
  \caption{$\nu$ Varying with Time}
  \includegraphics[width=\linewidth]{images/interpolated_nu_varying_with_time_IND.pdf}
  \label{fig:interpolated_nu_varying_with_time_IND}
\end{figure}

This shows us that the $\nu$ coincides with the actual of data of when the vaccination drive was first launched in India~\cite{UnicefCovidVaccine,PIBCovidVaccine,WikipediaCovidVaccine}. Therefore, using these values we finally, recompute $\beta_{optimal}$ and $\gamma_{optimal}$ by supplying them into the equations for the SIRV Model with time varying $\nu$.

\subsection{SIRV Model with Lockdown (Time-varying Vaccination Rate)}
Finally, we integrate the time-varying vaccination rate ($\nu$) into the SIR model that includes lockdown measures, resulting in the following set of equations, which represents our final model:
\begin{figure}[htbp!]
\begin{tikzpicture}[node distance=1cm, auto,
  >=Latex, 
  every node/.append style={align=center},
  int/.style={draw, minimum size=1cm}]
 \node [int] (S)             {$S$};
 \node [int, right=of S] (I) {$I$};
 \node [int, right=of I] (R) {$R$};
 \coordinate[right=of I] (out);
 \path[->] (S) edge node {$\lambda$} (I)
           (I) edge node {$\gamma$} (out)
           (S) edge[bend left=45] node[above] {$\nu(t)$} (R); % Arrow from S to R with text \nu(t)
\end{tikzpicture}
\end{figure}
\begin{equation}
  \frac{dS}{dt} = -\beta  (1 - s(t)/100)  \frac{S I}{N} - \nu(t) S
  \label{eq:S_with_lockdown_and_time_varying_nu}
\end{equation}
\begin{equation}
  \frac{dI}{dt} = \beta  (1 - s(t)/100) \frac{S I}{N} - \gamma I
  \label{eq:I_with_lockdown_and_time_varying_nu}
\end{equation}
\begin{equation}
  \frac{dR}{dt} = \gamma I + \nu(t) S
  \label{eq:R_with_lockdown_and_time_varying_nu}
\end{equation}
\begin{equation}
  \beta_{optimal} = 0.463
  \label{eq:beta_optimal_with_lockdown_and_time_varying_nu}
\end{equation}
\begin{equation}
  \gamma_{optimal} = 0.114
  \label{eq:gamma_optimal_with_lockdown_and_time_varying_nu}
\end{equation}
\begin{equation}
  \begin{split}
    \overbar{R_0} = 1.546 \\
    \widetilde{R_0} = 1.483 \\
    \text{Mode}(R_0) = 0.734 \\
    \sigma_{R_0} = 0.718 \\
    R_0 \in [0.150, 2.785]
  \end{split}
  \label{eq:r0_with_lockdown_and_time_varying_nu}
\end{equation}
\begin{equation}
  \textrm{loss\_SIR} = 29116762.926
  \label{eq:cost_SIR_with_lockdown_and_time_varying_nu}
\end{equation}
\begin{equation}
  \textrm{loss\_I} = 658537.443
  \label{eq:cost_I_with_lockdown_and_time_varying_nu}
\end{equation}
See \Crefrange{fig:SIR_model_IND}{fig:SIRV_model_with_lockdown_infections_time_varying_nu_IND} and \Crefrange{fig:comparing_costs_SIR_IND}{fig:comparing_costs_I_IND} to see how the different models compare against each other.

@@

The agent observes the percentage of the population that is susceptible, infected, recovered, and time varying data like the GDP, and previous actions taken to change the stringency, and $R_e$. Since Stable Baselines3 can support multiple inputs(time-series data, single data points and images) by using Dict Gym space. For data that varies with time (stringency, normalized GDP, $R_e$) we use a simple LSTM architecture~\cite{LSTM}. For other data like, the current proportion of the population that is susceptible, infected, and recovered we use a simple fully connected layer. The output from both these networks are concatenated and used by the reinforcement learning agent to train on.

>>>>

\title{SIR-RL: Reinforcement Learning for Optimized Policy Control during Epidemiological Outbreaks in Emerging Market and Developing Economies}
\author[1,*]{Maeghal Jain}
\author[1]{Ziya Uddin}

@@

In this paper we optimize the government policies regarding stringency as it controls both the spread of the disease and the economy. To model the epidemiological data~\cite{WorldometerCorona} we use the simple SIR model without vital dynamics~\cite{Hethcote1989, Hethcote2008, ALLEN2017128} as it is assumed that the timescale is small enough that it can be neglected~\cite{Cooper2020}. By lesioning the model, as opposed to proposing a new mathematical model with more specialized compartments to more accurately represent the actual environment~\cite{Bjrnstad2020, Mwalili2020}, we effectively model the disease progression. Our model (SIR with lockdown and time-varying vaccination rate) builds on the foundational SIR model, by accounting for the recovery reached through vaccination~\cite{Marinov2022,MaurciodeCarvalho2023,Thater,Turkyilmazoglu2022,YALADANDA2022101052} and the effects of lockdown~\cite{Hale2021,SIRLockdown,NBERw26981,NBERw26867}. Although the traditional SIR model is a valuable tool for understanding the spread of infectious diseases, it assumes that parameters like the transmission rate ($\beta$) and recovery rate ($\gamma$) are constant over time, which may not always be the case. In this paper, we propose a more sophisticated approach by introducing a time-dependent SIR model~\cite{TimeDependentSIR}, enabling us to account for the changing dynamics of the pandemic due to factors such as lockdowns and vaccination rates. This proposed model effectively addresses the real-world conditions acts as a solution that is both effective and extendable. However, the study has limitations; First, the deterministic SIR model (predecessor to our proposed model) fails to account for chance in disease spread and lacks confidence intervals on results and while stochastic models incorporate chance, they are typically more challenging to analyze than their deterministic counterparts~\cite{Hethcote2008}. Secondly, the underreporting of cases during the period selected by our study. Lastly, the reinforcement learning agent should be resistant to how the vaccination rate changes, and different values for $\beta$ and $\gamma$ -- keeping them the same scopes the environment for succumbing to wishful thinking which can be potentially dangerous. Therefore, before an actual deployment of the model, it would be a good measure to introduce stochasticity to these parameters($\beta$ and $\gamma$) and the vaccination rate.

@@

And, $\beta$ the effective transmission rate is defined as:
\begin{equation}
  \beta = p c
  \label{eq:beta_effective_transmission_rate}
\end{equation}
During an epidemic, the fundamental drivers of an epidemic growth is the rate of infection $\beta$ i.e., the average number of infections per infected case and the infectious period $1/\gamma$ i.e., the average period for which the infected case is infected for. Epidemics can only happen if the case is infectious enough for long enough and this defined by $R_0 = \beta / \gamma$. Here, $R_0$ is The average number of secondary infections caused by each infected case, in an otherwise fully susceptible population.

@@

To estimate the parameters $\beta$ and $\gamma$ for India, based on the data from May, 2020, to October, 2022, we simply define two cost functions \cref{eq:cost_SIR,eq:cost_I} to calibrate the model with the use of huber loss~\cite{huberloss}. We generally calibrate the model using \cref{eq:cost_SIR}. However, sometimes we must strike a balance between modelling all three population groups (susceptible, infected, and recovered) and focusing just on the infected group, which is most important because it drives the spread of the disease, therefore, in that case we consider the loss obtained from \cref{eq:cost_SIR,eq:cost_I} (see \cref{optimizing_window_length}, where a hyperparameter is selected keeping both these losses in mind).
\begin{equation}
  L_{\delta}(y, f(x)) = 
  \begin{cases}
    \frac{1}{2}{(y - f(x))^2} & \text{for } |y - f(x)| \leq \delta, \\
    \delta \cdot (|y - f(x)| - \frac{1}{2}\delta) & \text{otherwise.} \\
  \end{cases}
  \label{eq:huberloss}
\end{equation}
In the above equation, $y$ is the actual data and $f(x)$ is the prediction.
\begin{equation}
  \begin{split}
  \textrm{loss\_SIR} = \textrm{cost\_function\_SIR}(S, \hat{S}, I, \hat{I}, R, \hat{R}) \\
  = L_{\delta = 1}(S, \hat{S}) + L_{\delta = 1}(I, \hat{I}) + L_{\delta = 1}(R, \hat{R})
  \end{split}
  \label{eq:cost_SIR}
\end{equation}
\begin{equation}
  \textrm{loss\_I} = \textrm{cost\_function\_I}(I, \hat{I}) = L_{\delta = 1}(I, \hat{I})
  \label{eq:cost_I}
\end{equation}
Where, $S$ is the number of susceptible people and $\hat{S}$ is the predicted number of susceptible people, similarly, $I$ for infected and $\hat{I}$ for the predicted number of infected people, and $R$ for recovered. Using the equations \Crefrange{eq:S_without_lockdown}{eq:R_without_lockdown}, we minimize the cost function \cref{eq:cost_SIR_without_lockdown} using the Nelder-Mead method~\cite{Gao2010} to estimate the parameters like $\beta$, $\gamma$, and fit the model to actual data. The following parameters and loss is obtained:
\begin{equation}
  \beta_{optimal} = 0.042
  \label{eq:beta_optimal_without_lockdown}
\end{equation}
\begin{equation}
  \gamma_{optimal} = 0.024
  \label{eq:gamma_optimal_without_lockdown}
\end{equation}
\begin{equation}
  R_{0_{optimal}} = \frac{\beta_{optimal}}{\gamma_{optimal}} = 1.762
  \label{eq:r0_without_lockdown}
\end{equation}
\begin{equation}
  \textrm{loss\_SIR} = 85051490.533
  \label{eq:cost_SIR_without_lockdown}
\end{equation}
\begin{equation}
  \textrm{loss\_I} = 45187665.281
  \label{eq:cost_I_without_lockdown}
\end{equation}
See \cref{fig:SIR_model_IND} to see how the model compares with the actual data.

\subsection{SIR Model with Lockdown}
Now that, a simple SIR model has been established -- we need to model the effects of the stringency index (measure for the strictness of lockdown) on $\beta$ (the effective transmission rate). To do this, we say the flow of susceptibles not only depend on $\beta$ but also $s(t)$ the stringency index at time~\cite{NBERw26867,AlvarezLockdownSIR,SIRLockdown,SIRLockdown2,analytical_covid_lockdown_model}. The stringency index is a composite measure based on nine response indicators including school closures, workplace closures, and travel bans, rescaled to a value from 0 to 100 (100 = strictest)~\cite{owidcoronavirus}. This index simply records the strictness of government policies and does not measure or imply the appropriateness or effectiveness of a country's response i.e., a higher score does not necessarily mean that a country's response is "better" than others lower on the index.

To define the new time-varying beta that is dependent on the current stringency index, the following equations have been formulated:
\begin{equation}
  \frac{dS}{dt} = -\beta (1 - s(t)/100) \frac{S I}{N}
  \label{eq:S_with_lockdown}
\end{equation}
\begin{equation}
  \frac{dI}{dt} = \beta (1 - s(t)/100) \frac{S I}{N} - \gamma I
  \label{eq:I_with_lockdown}
\end{equation}
\begin{equation}
  \frac{dR}{dt} = \gamma I
  \label{eq:R_with_lockdown}
\end{equation}
Where, $s(t)$ is the stringency index at time $t$ and is scaled down by a factor of 100 to normalize it and bring it in the range $s(t)/100 \in [0, 1]$. Multiplying the rate of flow from $S$ to $I$ compartment with $1 - s(t)/100$ allows us to account for the effect that stringency has on the disease progression. A higher stringency index can theoretically, stop the flow from the susceptible population to the infected population entirely. Optimizing these equations with \cref{eq:cost_SIR} using the Nelder-Mead method, we get the following parameters and loss:
\begin{equation}
  \beta_{optimal} = 0.401
  \label{eq:beta_optimal_with_lockdown}
\end{equation}
\begin{equation}
  \gamma_{optimal} = 0.090
  \label{eq:gamma_optimal_with_lockdown}
\end{equation}
\begin{equation}
  \begin{split}
    \overbar{R_0} = 1.693 \quad \text{(Mean)} \\
    \widetilde{R_0} = 1.624 \quad \text{(Median)} \\
    \text{Mode}(R_0) = 0.804 \quad \text{(Mode)} \\
    \sigma_{R_0} = 0.786 \quad \text{(Standard Deviation)} \\
    R_0 \in [0.16467, 3.0497] \quad \text{(Range)}
  \end{split}
  \label{eq:r0_with_lockdown}
\end{equation}
\begin{equation}
  \textrm{loss\_SIR} = 98438821.456
  \label{eq:cost_SIR_with_lockdown}
\end{equation}
\begin{equation}
  \textrm{loss\_I} = 11345389.686
  \label{eq:cost_I_with_lockdown}
\end{equation}
\subsection{SIR Model with Lockdown and Vaccination}
Lastly, an additional flow from the susceptible to recovered population can be shown by adding a vaccination rate $\nu$ in the model. 
\begin{equation}
  \frac{dS}{dt} = -\beta (1 - s(t)/100) \frac{S I}{N} - \nu S
  \label{eq:S_with_lockdown_and_nu}
\end{equation}
\begin{equation}
  \frac{dI}{dt} = \beta (1 - s(t)/100) \frac{S I}{N} - \gamma I
  \label{eq:I_with_lockdown_and_nu}
\end{equation}
\begin{equation}
  \frac{dR}{dt} = \gamma I + \nu S
  \label{eq:R_with_lockdown_and_nu}
\end{equation}
Optimizing these equations with \cref{eq:cost_SIR} using the Nelder-Mead method:
\begin{equation}
  \beta_{optimal} = 0.409
  \label{eq:beta_optimal_with_lockdown_and_nu}
\end{equation}
\begin{equation}
  \gamma_{optimal} = 0.092
  \label{eq:gamma_optimal_with_lockdown_and_nu}
\end{equation}
\begin{equation}
  \nu_{optimal} = 2.904 \times 10^{-5}
  \label{eq:nu_optimal_with_lockdown_and_nu}
\end{equation}
\begin{equation}
  \begin{split}
    \overbar{R_0} = 1.691 \\
    \widetilde{R_0} = 1.623 \\
    \text{Mode}(R_0) = 0.803 \\
    \sigma_{R_0} = 0.785 \\
    R_0 \in [0.165, 3.047]
  \end{split}
  \label{eq:r0_with_lockdown_and_nu}
\end{equation}
\begin{equation}
  \textrm{loss\_SIR} = 94636860.384
  \label{eq:cost_SIR_with_lockdown_and_nu}
\end{equation}
\begin{equation}
  \textrm{loss\_I} = 10840360.995
  \label{eq:cost_I_with_lockdown_and_nu}
\end{equation}

\subsection{Optimizing Window Length for Time-varying Vaccination Rate}\label{optimizing_window_length}
However, as observed by the value of $\nu_{optimal}$ from \cref{eq:nu_optimal_with_lockdown_and_nu} which is almost negligible and the overestimation of infected individuals in \cref{fig:SIR_model_with_lockdown_with_vaccination_infections_IND} suggests that $\nu$ might be varying with time. This suggests to accurately estimate the infected population a time-varying vaccination rate should be used as the transition from susceptibility to direct recovery fluctuates with time~\cite{Liang2021,Marinov2022}. Therefore, using the $\beta_{optimal}$ and $\gamma_{optimal}$ from \cref{eq:beta_optimal_with_lockdown_and_nu} and \cref{eq:gamma_optimal_with_lockdown_and_nu}, we first find the optimal window length~\cite{Liao2020} i.e., a window length that results in the least loss from \cref{eq:cost_SIR,eq:cost_I}. Using different window lengths ($\textrm{window\_length} = 5, 10, 15 \dots 40, 45, 50$), we estimate value of $\nu$ given a sub-interval of $[{start}, {start} + {window\_length}]$ then for the next time-step $[{start} + {window\_length}, {start} + {window\_length} + {window\_length}]$ we again estimate $\nu$ -- so on and so forth. It is crucial to note that the variable $\nu$ is constrained to be a positive integer, reflecting the inherent one-way nature of vaccination: individuals can only receive vaccinations, not return them.
\begin{equation}
  \frac{dS}{dt} = -\beta_{optimal}  (1 - s(t)/100)  \frac{S I}{N} - \nu S
  \label{eq:S_with_lockdown_and_nu_calc_nu}
\end{equation}
\begin{equation}
  \frac{dI}{dt} = \beta_{optimal}  (1 - s(t)/100) \frac{S I}{N} - \gamma_{optimal} I
  \label{eq:I_with_lockdown_and_nu_calc_nu}
\end{equation}
\begin{equation}
  \frac{dR}{dt} = \gamma_{optimal} I + \nu S
  \label{eq:R_with_lockdown_and_nu_calc_nu}
\end{equation}
This model gives shows how $\nu$ varies with time. Since, the window length acts like a hyperparameter, we can try different window lengths in order to evaluate which would give us the best results. Therefore, the below graphs show us the loss using \cref{eq:cost_SIR,eq:cost_I}.

\begin{figure}[htbp!]
  \begin{subfigure}[t]{0.48\textwidth}
    \centering
    \includegraphics[width=\linewidth]{images/window_length_loss_SIR_IND.pdf}
    \caption{Loss for Different Window Lengths for Susceptible, Infected and Recovered Population}
    \label{fig:window_length_loss_SIR_IND}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.48\textwidth}
    \centering
    \includegraphics[width=\linewidth]{images/window_length_loss_I_IND.pdf}
    \caption{Loss for Different Window Lengths for Infected Population}
    \label{fig:window_length_loss_I_IND}
  \end{subfigure}
  \caption{\textbf{Loss for Different Window Lengths}. Here we change the window length to see which window length yields the best loss in both cases, while predicting the susceptible, infected and recovered populations or just the infected population.}
\end{figure}

The following results in \Crefrange{fig:window_length_loss_SIR_IND}{fig:window_length_loss_I_IND}, indicate that a window length of $10$ days yields the least overall loss for all three population groups. However, this window length results in a poor approximation for just the infected group, which is crucial for accurately modelling the spread of the disease. Consequently, we have decided to use a window length of $15$ days, which provides a more accurate approximation for the infected population while still maintaining reasonable loss for the other groups.

\begin{figure}[htbp!]
  \centering
  \includegraphics[width=\linewidth]{images/interpolated_nu_varying_with_time_IND.pdf}
  \label{fig:interpolated_nu_varying_with_time_IND}
  \caption{\textbf{$\nu$ Varying with Time.} This shows how the vaccination rate $\nu$ varies with time and when the vaccination drive was launched in India.}
\end{figure}

This shows us that the $\nu$ coincides with the actual of data of when the vaccination drive was first launched in India~\cite{UnicefCovidVaccine,PIBCovidVaccine,WikipediaCovidVaccine}. Therefore, using these values we finally, recompute $\beta_{optimal}$ and $\gamma_{optimal}$ by supplying them into the equations for the SIR Model with lockdown and time-varying $\nu$.

\subsection{SIR Model with Lockdown and Time-varying Vaccination Rate}
Finally, we integrate the time-varying vaccination rate ($\nu$) into the SIR model that includes lockdown measures, resulting in the following set of equations, which represents our final model:
\begin{figure}[htbp!]
\begin{tikzpicture}[node distance=1cm, auto,
  >=Latex, 
  every node/.append style={align=center},
  int/.style={draw, minimum size=1cm}]
 \node [int] (S)             {$S$};
 \node [int, right=of S] (I) {$I$};
 \node [int, right=of I] (R) {$R$};
 \coordinate[right=of I] (out);
 \path[->] (S) edge node {$\lambda$} (I)
           (I) edge node {$\gamma$} (out)
           (S) edge[bend left=45] node[above] {$\nu(t)$} (R); % Arrow from S to R with text \nu(t)
\end{tikzpicture}
\end{figure}
\begin{equation}
  \frac{dS}{dt} = -\beta  (1 - s(t)/100)  \frac{S I}{N} - \nu(t) S
  \label{eq:S_with_lockdown_and_time_varying_nu}
\end{equation}
\begin{equation}
  \frac{dI}{dt} = \beta  (1 - s(t)/100) \frac{S I}{N} - \gamma I
  \label{eq:I_with_lockdown_and_time_varying_nu}
\end{equation}
\begin{equation}
  \frac{dR}{dt} = \gamma I + \nu(t) S
  \label{eq:R_with_lockdown_and_time_varying_nu}
\end{equation}
\begin{equation}
  \beta_{optimal} = 0.463
  \label{eq:beta_optimal_with_lockdown_and_time_varying_nu}
\end{equation}
\begin{equation}
  \gamma_{optimal} = 0.114
  \label{eq:gamma_optimal_with_lockdown_and_time_varying_nu}
\end{equation}
\begin{equation}
  \begin{split}
    \overbar{R_0} = 1.546 \\
    \widetilde{R_0} = 1.483 \\
    \text{Mode}(R_0) = 0.734 \\
    \sigma_{R_0} = 0.718 \\
    R_0 \in [0.150, 2.785]
  \end{split}
  \label{eq:r0_with_lockdown_and_time_varying_nu}
\end{equation}
\begin{equation}
  \textrm{loss\_SIR} = 29116762.926
  \label{eq:cost_SIR_with_lockdown_and_time_varying_nu}
\end{equation}
\begin{equation}
  \textrm{loss\_I} = 658537.443
  \label{eq:cost_I_with_lockdown_and_time_varying_nu}
\end{equation}
See \Crefrange{fig:SIR_model_IND}{fig:SIR_model_with_lockdown_with_vaccination_infections_time_varying_nu_IND} and \Crefrange{fig:comparing_costs_SIR_IND}{fig:comparing_costs_I_IND} to see how the different models compare against each other.

@@

The agent observes the percentage of the population that is susceptible, infected, recovered, and time-varying data like the GDP, and previous actions taken to change the stringency, and $R_e$. Since Stable Baselines3 can support multiple inputs(time-series data, single data points and images) by using Dict Gym space. For data that varies with time (stringency, normalized GDP, $R_e$) we use a simple LSTM architecture~\cite{LSTM}. For other data like, the current proportion of the population that is susceptible, infected, and recovered we use a simple fully connected layer. The output from both these networks are concatenated and used by the reinforcement learning agent to train on.
